Verification of Stencil Computations through Spatial Specifications

### 2016-07-03 paper.tex
added details on successful parsing and explanation of failures

\item Compute all permutations on the column vectors in a span, \ie{},
  $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
for a permutation $\pi$. For each permutation function $\pi^n_i$
(the $i$-th permutation for vectors of size $n$, we pair the
permutation function with the set of permuted spans so that
the spans can be unpermuted later.
%
\[
P(v) = \bigcup_{i \in !n} (\pi^n_{i} , \; \{[\pi^n_i
\vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
\leftarrow N(v)\}
\]
%
For our example, this yields:
%
\begin{align*}
P(\texttt{a}) =
\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0}
\stwo{1}{0}{1}{0}
\stwo{-1}{0}{-1}{0}
\stwo{0}{1}{0}{1}
\stwo{0}{-1}{0}{-1} \})
\\
&(\pi^2_2, \{
 \stwo{0}{0}{0}{0}
 \stwo{0}{1}{0}{1}
 \stwo{0}{-1}{0}{-1}
 \stwo{1}{0}{1}{0}
 \stwo{-1}{0}{-1}{0}\})\}
\end{align*}
%
where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
permutation flips the order of the two elements in the column
vectors.

\item Sort each permutation set into an ordered list, based on the ordering:
\[
\vect{x} \leq \vect{y} = \exists i \, . \; \vect{x}_i \leq \vect{y}_i
\; \wedge \; \vect{x}_{i:n} = \vect{y}_{i:n}
\]
%
This yields:
%%
\begin{align*}
\{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{-1}{0}
\stwo{0}{0}{0}{0}
\stwo{1}{0}{1}{0}
\stwo{0}{1}{0}{1}] )
\\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{-1}{0}
\stwo{0}{0}{0}{0}
\stwo{1}{0}{1}{0}
\stwo{0}{1}{0}{1}])\}
\end{align*}
%%
\item Fold each list pairwise by the following partial operation
 $\bullet$ which coalescees contiguous regions:
%
\begin{align*}
& (\vect{x}^L,\vect{x}^U) \bullet (\vect{y}^L,\vect{y}^U) \\
= &
\begin{cases}
(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
(\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) = (\vect{y}^L_{1:n}, \vect{y}^U_{1:n}) \\
\bot  & \textit{otherwise}
\end{cases}
\end{align*}
For our example, this yields:
%
\begin{align*}
Q(v) = \{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1},
]) \\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
%
\item Unpermute and union together, \ie{},
%
\[
U(v) = \bigcup \{[\pi \vect{y}^L, \pi \vect{y}^U]
 \mid [\vect{y}^L, \vect{y}^U] \leftarrow S, (\pi, S) \leftarrow Q(v)\}
\]
Which yields:
%
\begin{align*}
U(\texttt{a}) =
\{\stwo{0}{-1}{0}{-1}
\stwo{-1}{0}{1}{0}
\stwo{0}{1}{0}{1}
\stwo{-1}{0}{-1}{0}
\stwo{0}{-1}{0}{1}
\stwo{1}{0}{1}{0}\}
\end{align*}
%
\item Finally, filter by the region containment predicate $\containedin$, that
  is, if any region is contained within another then remove the
  smaller. The $\containedin$ predicate is defined:
%
\begin{align*}
& (\vect{x}^L, \vect{x}^U) \containedin (\vect{y}^L, \vect{y}^U) = \\
& \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
  \wedge (\vect{x}^L_{1:n}, \vect{x}^U_{1:n}) \containedin
  (\vect{y}^L_{1:n}, \vect{y}^U_{1:n})
\end{align*}
For our example this then yields the final result of:
\begin{align*}
\textsf{regions}(M)
= \{\stwo{-1}{0}{1}{0} \stwo{0}{-1}{0}{1}\}
\end{align*}
since $\stwo{0}{1}{0}{1} \sqsubseteq \stwo{0}{-1}{0}{1}$
and $\stwo{1}{0}{1}{0} \sqsubseteq \stwo{-1}{0}{1}{0}$.
\end{enumerate}
For our example, applying \textsf{regions} again yields the same
result, hence we have reached a fixed point.

@@

\section{Evaluation}
\label{sec:evaluation}
We now present an evaluation of the stencil language. We first confirm
that stencils are common in scientific code and further that our
stencil language can express a great many of these.

\subsection{Software Corpus}
\begin{figure}
\begin{tabular}{|l|l|l|}
\hline
Package       & Files & Lines of code \\
\hline
UM            & \umFiles  & \umLoC       \\
E3MG          & \ethreemgeaFiles   & \ethreemgeaLoC        \\
BLAS          & \blasFiles   & \blasLoC        \\
Hybrid4       & \hybridfourFiles    & \hybridfourLoC          \\
GEOS-Chem     & \geoschemFiles   & \geoschemLoC       \\
Navier        & \navierFiles     & \navierLoC           \\
CP            & \computationalphysicstwoFiles    & \computationalphysicstwoLoC           \\
\hline
\textbf{Total}& \overallFiles  & \overallLoC     \\
\hline 
\end{tabular}
\caption{Summary of software packages used for evaluation\label{fig:corpus}}
\end{figure}

Figure \ref{fig:corpus} shows summary statistics of the software packages used in our evaluation. In total we analysed 1.1 million lines of code from 7 packages.

>>>>

\item Compute all permutations on the column vectors in a span, \ie{},
  $[\vect{x}^L \vect{x}^U] \mapsto [\pi\vect{x}^L \, \pi\vect{x}^U]$
for a permutation $\pi$. For each permutation function $\pi^n_i$
(the $i$-th permutation for vectors of size $n$, we pair the
permutation function with the set of permuted spans so that
the spans can be unpermuted later.
%
\[
P(v) = \bigcup_{i \in n!} (\pi^n_{i} , \; \{[\pi^n_i
\vect{x}^L \; \pi^n_i\vect{x}^U] \, \mid \, [\vect{x}^L \; \vect{x}^U]
\leftarrow N(v)\}
\]
%
For our example, this yields:
%
\begin{align*}
P(\texttt{a}) =
\{&(\pi^2_1, \{ \stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{-1}{0}{-1}{0},
\stwo{0}{1}{0}{1},
\stwo{0}{-1}{0}{-1} \})
\\
&(\pi^2_2, \{
 \stwo{0}{0}{0}{0},
 \stwo{0}{1}{0}{1},
 \stwo{0}{-1}{0}{-1},
 \stwo{1}{0}{1}{0},
 \stwo{-1}{0}{-1}{0}\})\}
\end{align*}
%
where $\pi^2_1$ is the identity permutation and $\pi^2_2$ is the
permutation flips the order of the two elements in the column
vectors.

\item Sort each permutation set into an ordered list, based on the ordering:
\begin{equation*}
  \vect{x} \leq \vect{y} =
      \exists i \, . \; \vect{x}_{i,1} \leq \vect{y}_{i,1} \; \wedge \;
        (i = n \vee \vect{x}_{i+1:n} = \vect{y}_{i+1:n})
\end{equation*}
%
This yields:
%%
\begin{align*}
\{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{0}{1}{0}{1}] )
\\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{0}{0}{0},
\stwo{1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
%%
\item Fold each list pairwise by the following partial operation
 $\bullet$ which coalescees contiguous regions:
%
\begin{align*}
& (\vect{x}^L,\vect{x}^U) \bullet (\vect{y}^L,\vect{y}^U) \\
= &
\begin{cases}
(\vect{x}^L, \vect{y}^U) & \vect{x}^U_1 + 1 = \vect{y}^L_1 \; \wedge \;
\vect{x}_{2:n} = \vect{y}_{2:n}) \\
\bot  & \textit{otherwise}
\end{cases}
\end{align*}
For our example, this yields:
%
\begin{align*}
Q(v) = \{&(\pi^2_1, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}
]) \\
&(\pi^2_2, [
\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1}])\}
\end{align*}
%
\item Unpermute and union together, \ie{},
%
\[
U(v) = \bigcup \{[\pi \vect{x}^L, \pi \vect{x}^U]
 \mid [\vect{x}^L, \vect{x}^U] \leftarrow S, (\pi, S) \leftarrow Q(v)\}
\]
Which yields:
%
\begin{align*}
U(\texttt{a}) =
\{\stwo{0}{-1}{0}{-1},
\stwo{-1}{0}{1}{0},
\stwo{0}{1}{0}{1},
\stwo{-1}{0}{-1}{0},
\stwo{0}{-1}{0}{1},
\stwo{1}{0}{1}{0}\}
\end{align*}
%
\item Finally, filter by the region containment predicate $\containedin$, that
  is, if any region is contained within another then remove the
  smaller. The $\containedin$ predicate is defined:
%
\begin{align*}
& (\vect{x}^L, \vect{x}^U) \containedin (\vect{y}^L, \vect{y}^U) = \\
& \vect{y}^L_1 \leq \vect{x}^L_1 \wedge \vect{x}^U_1 \leq \vect{y}^U_1
  \wedge (\vect{x}^L_{2:n}, \vect{x}^U_{2:n}) \containedin
  (\vect{y}^L_{2:n}, \vect{y}^U_{2:n})
\end{align*}
For our example this then yields the final result of:
\begin{align*}
\textsf{regions}(M)
= \{\stwo{-1}{0}{1}{0} \stwo{0}{-1}{0}{1}\}
\end{align*}
since $\stwo{0}{1}{0}{1} \sqsubseteq \stwo{0}{-1}{0}{1}$
and $\stwo{1}{0}{1}{0} \sqsubseteq \stwo{-1}{0}{1}{0}$.
\end{enumerate}
For our example, applying \textsf{regions} again yields the same
result, hence we have reached a fixed point.

@@

\section{Evaluation}
\label{sec:evaluation}
We evaluated our stencil language on a corpus of scientific code. We first give an overview of the corpus itself before presenting our results.

\subsection{Software Corpus}
\begin{figure}
\begin{tabular}{|l|l|l|l|}
\hline
Package       & Files & Lines of code & Files analyzed \\
\hline
UM            & \umFiles & \umLoC & \umparseOk  \\
E3MG          & \ethreemgeaFiles   & \ethreemgeaLoC & \ethreemgeaparseOk  \\
BLAS          & \blasFiles   & \blasLoC & \blasparseOk        \\
Hybrid4       & \hybridfourFiles    & \hybridfourLoC & \hybridfourparseOk         \\
GEOS-Chem     & \geoschemFiles   & \geoschemLoC & \geoschemparseOk       \\
Navier        & \navierFiles     & \navierLoC & \navierparseOk           \\
CP            & \computationalphysicstwoFiles    & \computationalphysicstwoLoC & \computationalphysicstwoparseOk           \\
\hline
\textbf{Total}& \overallFiles  & \overallLoC   & \overallparseOk  \\
\hline 
\end{tabular}
\caption{Summary of software packages used for evaluation\label{fig:corpus}}
\end{figure}

Figure \ref{fig:corpus} shows summary statistics of the software packages used in our evaluation, all of which are written in Fortran 90 or Fortran 77. In total we analysed 1.1 million lines of code from 7 packages. The ``Files analyzed'' column shows the number of files in each corpus that we were able to analyze with Camfort. The most common reasons for Camfort rejecting a file were either use of a C-preprocessor, or illegal use of language features from a more modern Fortran variant.
